Investigation on Parallel Training of Deep Neural Network using Model Averaging

### 2015-06-25 paper.tex
more edits

\section{Data parallelization and Model Averging}
SGD is a popular method for DNN training. Even though neural network training objectives are usually non-convex, 
mini-batch SGD has been shown to be effective for optimizing the objective\cite{seide2011conversational}. 
Roughly speaking, a bigger minibatch size gives a better estimate of the gradient, resulting in a better the converge rate. 
Thus, a straight forward idea for parallellization would be distributing the gradient computation to different computing
nodes. In each step, gradients of minibatches on different nodes are reduced to a single node, averaged and then used to 
update models in each node. This method can compute the gradient accurately, but it requires heavy communication between nodes.
Also, it is shown that increasing minibatch size does not always benefit model training\cite{seide2011conversational}, 
especially in early stage of model training.

@@

\subsection{Switchboard Results}
scaling factor
\begin{figure}[htb]
  \centering
  \includegraphics[width=0.5\textwidth]{scaling.png}
  \caption{Scaling factor and speedup factor v.s. number of gpus}
  \label{fig:scaling}
\end{figure}

percentage of time spent on communication
\begin{figure}[htb]
  \centering
  \includegraphics[width=0.5\textwidth]{percent.png}
  \caption{Percentage of time on MPI communication v.s. number of gpus}
  \label{fig:percent}
\end{figure}

\begin{table}
  \centering
  \begin{tabular}{c|c|c|c|c|c|c}
    \hline
     Nodes  & 1    & 2    & 4    & 8    & 16 & 32\\
    \hline
Allreduce &    14.7 & -- & -- & 15.1 & 15.1   & 15.2\\
    \hline
Butterfly &    14.7 & -- & -- & --   & 15.6   & 16.0\\
    \hline
  \end{tabular}
  \caption{Comparison of WERs using different number of GPUs }
  \label{tab:wer}
\end{table}

It is shown that this is

\subsection{Initialization Matters}
\label{sec:init}
Compare random init and dbn init.

\begin{table}
  \centering
  \begin{tabular}{c|c|c|c|c}
    \hline
    & \multicolumn{2}{c|}{SWB}  & \multicolumn{2}{c}{CallHome} \\
    \hline
    Nodes          &  1     & 32     &   1    & 32\\
    \hline
    random init     & RR    & 16.4   &  RR    & 28.8 \\
    \hline
    rbm pretraining & 14.7  & 15.2   & 26.8   & 27.1 \\
    \hline
  \end{tabular}
  \caption{Comparing rbm pretraining with random initialization}
  \label{tab:init}
\end{table}

\subsection{Averaging frequency}
Averaging frequency here is defined as the number of minibatch-SGD performed per model averaging.
Due to the limitation of computing resource, we only did preliminary experiments on this. Minibatch size of 1024 
is set as default, and we compare averaging frequency of 10 and 20. It is shown that an averaging frequency of 10 gives
slight worse speedup but a better decoding WER. The tradeoff between lower averaging frequency (i.e. better speedup) and
better training accuracy is within expectation in that frequent model averaging means steady gradient estimation.

\begin{table}
  \centering
  \begin{tabular}{c|c|c|c}
    \hline
    frequency   & Speedup   & SWB   & CallHome \\
    \hline
    baseline    &   --      & 14.7  & 26.8\\
    \hline
    10          & 9.32      & 15.1  & 27.0 \\
    \hline
    20          & 10.07     & 15.8  & 28.0 \\
    \hline
  \end{tabular}
  \caption{Comparing different averaging frequencies}
  \label{tab:init}
\end{table}

\subsection{Minibatch Size}
\begin{table}
  \centering
  \begin{tabular}{c|c|c|c|c|c}
    \hline
    & \multicolumn{2}{c|}{SWB}    & \multicolumn{2}{c}{CallHome} & Speedup\\
    \hline
    nodes       &  1    &   16    &    1       &   16   &    \\
    \hline
    256         &  --      & 15.1  & 27.0     &   & \\
    \hline
    1024        & 9.32     & 15.1  & 27.0     &   & \\
    \hline
  \end{tabular}
  \caption{Comparing different minibatch size}
  \label{tab:init}
\end{table}

\subsection{Learing Rate Schedule}
Initial learning rate is increased in porportion to number of threads in this setup. The reason for this is straight forward:
Assume we have $n$ minibatches of data for model training. When the model is trained using single thread, it gets updated $n$
times. When data is distributed to $m$ machines, then each model gets updated $n/m$ times. Since the effect of model averaging 
is mostly aggregating knowledge learnt from different data partition, the absolute change of model shall be compensated by 
$m$ times.

@@

\begin{table}
  \centering
  \begin{tabular}{c|c|c|c|c}
    \hline
    & \multicolumn{2}{c|}{SWB}  & \multicolumn{2}{c}{CallHome} \\
    \hline
    Nodes       &  1      & 16    &  1    & 16 \\
    \hline
    Newbob      & 14.9    & 15.4  & 26.6  & 27.2 \\
    \hline
    exponential & 14.7    & 15.1  & 26.8  & 27.0 \\
    \hline
  \end{tabular}
  \caption{Comparing learning rate schedule}
  \label{tab:init}
\end{table}

\subsection{Online NG-SGD Matters}
\begin{table}
  \centering
  \begin{tabular}{c|c|c|c|c}
    \hline
    & \multicolumn{2}{c|}{SWB}  & \multicolumn{2}{c}{CallHome} \\
    \hline
     Nodes  &  1    & 16        &   1   &  16 \\
    \hline
    SGD     & 14.9    & 16.3    &  26.9 & 28.3 \\
    \hline
    NG-SGD  & 14.7    & 15.1    &  26.8 & 27.0  \\
    \hline
  \end{tabular}
  \caption{Comparing NG-SGD and naive SGD}
  \label{tab:ngsgd}
\end{table}

\section{Conclusion and Future Work}
Neural network training can be efficiently speed up by using model averaging techniques and NG-SGD. 
Butterfly-mixing and ring reduction can reduce communication time a lot, but the models may not converge as well as 
Allreduce when the number of jobs goes up to 16.

Further improvement might be achieved if parallel training runs on top of serial training initialization.
CUDA aware MPI, learning rate schedule. Averaging frequency. Theory on convergence.

@@

% References should be produced using the bibtex program from suitable
% BiBTeX files (here: strings, refs, manuals). The IEEEbib.bst bibliography
% style file from IEEE produces unsorted bibliography list.
% -------------------------------------------------------------------------
\bibliographystyle{IEEEbib}
\bibliography{paper}

>>>>

\section{Relationship to Prior Works}
To avoid confusion, we should mention that Kaldi\cite{kaldi11} contains two neural network recipes. The first implementation
\footnote{Location in code: src/\{nnet,nnetbin\}} is described in \cite{vesely2013sequence} which supports Restricted Boltzmann Machine 
pretraining \cite{hinton2006fast} and sequence-discriminative training \cite{povey2008boosted}. It uses single GPU 
for SGD training. The second implementation \footnote{Location in code: src/{nnet2,nnet2bin}}
\cite{zhang2014improving} was originally designed to support parallel training on multiple CPUs. Now it also supports 
multiple GPUs for training using model averaging. By default, it uses layer-wise discriminative pretraining.

Our work extends the first implementation so that it can utilize multiple GPUs using model averaging. We use MPI 
in implementation, so file I/O is avoided during model averaging. This allows us to perform model averaging much
more frequently.

\section{Data parallelization and Model Averging}
SGD is a popular method for DNN training. Even though neural network training objectives are usually non-convex, 
mini-batch SGD has been shown to be effective for optimizing the objective\cite{seide2011conversational}. 
Roughly speaking, a bigger minibatch size gives a better estimate of the gradient, resulting in a better the converge rate. 
Thus, a straight forward idea for parallellization would be distributing the gradient computation to different computing
nodes. In each step, gradients of minibatches on different nodes are reduced to a single node, averaged and then used to 
update models in each node. This method, i.e. gradient averaging, can compute the gradient accurately, but it requires 
heavy communication between nodes.
Also, it is shown that increasing minibatch size does not always benefit model training\cite{seide2011conversational}, 
especially in early stage of model training.

@@

\subsection{Switchboard Results}
Fig.~\ref{fig:scaling} shows the scaling factor and speedup plot for model averaging experiments. As is shown in the graph,
a speedup of 17 could be achieved when 32 GPUs are used.
\begin{figure}[htb]
  \centering
  \includegraphics[width=0.5\textwidth]{scaling.png}
  \caption{Scaling factor and speedup factor v.s. number of gpus}
  \label{fig:scaling}
\end{figure}
Table~\ref{tab:wer} shows the main decoding results for DNNs trained using different number of GPUs. In general, decoding
results of DNNs trained model averaging degrades 0.3\~0.4 WER, depending on the number of GPUs used.
\begin{table}
  \centering
  \begin{tabular}{c|c|c|c|c|c|c}
    \hline
     Nodes  & 1    & 2    & 4   & 8      & 16     & 32\\
    \hline
    SWB     & 14.7 & 14.9 & --  & 15.1   & 15.1   & 15.2\\
    CallHM  & 26.8 & 26.8 & --  & 27.4   & 27.0   & 27.1 \\
    \hline
  \end{tabular}
  \caption{Comparison of WERs using different number of GPUs}
  \label{tab:wer}
\end{table}

\subsection{Initialization Matters}
\label{sec:init}
Table~\ref{tab:init} compares random initialization with Restricted Boltzmann Machine (RBM) based initialization.

\begin{table}
  \centering
  \begin{tabular}{c|c|c|c|c}
    \hline
    & \multicolumn{2}{c|}{SWB}  & \multicolumn{2}{c}{CallHome} \\
    \hline
    Nodes          &  1     & 32     &   1    & 32\\
    \hline
    random init     & --    & 16.4   &  --    & 28.8 \\
    \hline
    RBM pretraining & 14.7  & 15.2   & 26.8   & 27.1 \\
    \hline
  \end{tabular}
  \caption{Comparing RBM pretraining with random initialization}
  \label{tab:init}
\end{table}
As we can see in the table, random initialization is worse than RBM pretraining based DNN by xx in single GPU case.
In model averaging setup, random initialization becomes even worse -- a xx point more degradation on WER.

\subsection{Averaging frequency}
Averaging frequency here is defined as the number of minibatch-SGD performed per model averaging.
Due to the limitation of computing resource, we only did preliminary experiments on this. Minibatch size of 1024 
is set as default, and we compare averaging frequency of 10 and 20. It is shown in Table~\ref{tab:freq} that an averaging 
frequency of 10 gives slight worse speedup but a better decoding WER. The tradeoff between lower averaging frequency 
(i.e. better speedup) and better training accuracy is within expectation in that frequent model averaging means 
steady gradient estimation.

\begin{table}
  \centering
  \begin{tabular}{c|c|c|c}
    \hline
    frequency   & Speedup   & SWB   & CallHome \\
    \hline
    baseline    &   --      & 14.7  & 26.8\\
    \hline
    10          & 9.32      & 15.1  & 27.0 \\
    \hline
    20          & 10.07     & 15.8  & 28.0 \\
    \hline
  \end{tabular}
  \caption{Comparing different averaging frequencies}
  \label{tab:freq}
\end{table}

\subsection{Minibatch Size}
Table~\ref{tab:mbsize} compares two different minibatch size in model averaging setup. It is shown that.
\begin{table}
  \centering
  \begin{tabular}{c|c|c|c|c|c}
    \hline
    & \multicolumn{2}{c|}{SWB}    & \multicolumn{2}{c|}{CallHome} & Speedup\\
    \hline
    nodes       &  1    &   16    &    1     &   16   &    \\
    \hline
    256         &  --      &      &  --      &        & \\
    \hline
    1024        & 9.32     & 15.1 & 27.0     &        & \\
    \hline
  \end{tabular}
  \caption{Comparing different minibatch size}
  \label{tab:mbsize}
\end{table}

\subsection{Learing Rate Schedule}
Initial learning rate is increased in porportion to number of threads in model averaging setup. The reason for this is straight 
forward:
Assume we have $n$ minibatches of data for model training. When the model is trained using single thread, it gets updated $n$
times. When data is distributed to $m$ machines, then each model gets updated $n/m$ times. Since the effect of model averaging 
is mostly aggregating knowledge learnt from different data partition, the absolute change of model shall be compensated by 
$m$ times.

@@

As is shown in Table~\ref{tab:lr}, exponential learning rate gives slight better decoding results than Newbob.
\begin{table}
  \centering
  \begin{tabular}{c|c|c|c|c}
    \hline
    & \multicolumn{2}{c|}{SWB}  & \multicolumn{2}{c}{CallHome} \\
    \hline
    Nodes       &  1      & 16    &  1    & 16 \\
    \hline
    Newbob      & 14.9    & 15.4  & 26.6  & 27.2 \\
    \hline
    exponential & 14.7    & 15.1  & 26.8  & 27.0 \\
    \hline
  \end{tabular}
  \caption{Comparing learning rate schedule}
  \label{tab:lr}
\end{table}

\subsection{Online NG-SGD Matters}
Table~\ref{tab:ngsgd} compares plain SGD with NG-SGD in model averaging mode, and it shows NG-SGD with 
\begin{table}
  \centering
  \begin{tabular}{c|c|c|c|c}
    \hline
    & \multicolumn{2}{c|}{SWB}  & \multicolumn{2}{c}{CallHome} \\
    \hline
     Nodes  &  1    & 16        &   1   &  16 \\
    \hline
    SGD     & 14.9    & 16.3    &  26.9 & 28.3 \\
    \hline
    NG-SGD  & 14.7    & 15.1    &  26.8 & 27.0  \\
    \hline
  \end{tabular}
  \caption{Comparing NG-SGD and naive SGD}
  \label{tab:ngsgd}
\end{table}

\section{Conclusion and Future Work}
In this paper, we show that neural network training can be efficiently speeded up using model averaging. 
on a 300h Switchboard dataset, a 9.3x / 17x speedup could be achieved using 16 / 32 GPUs respectively, 
with limited decoding accuracy loss. We also show that model averaging benefits a lot from NG-SGD and RBM based pretraining.
Preliminary experiments on minibatch size, averaging frequency and learning rate schedules are also presented.

Further accuracy improvement might be achieved if parallel training runs on top of serial training initialization.
It would be interesting to see if sequence-discriminative training combines well with model averaging.
Speedup factor could be further improved if CUDA aware MPI is used. Theory on convergence using model averaging is 
to be explored, which might be useful for guiding future development.

@@

% References should be produced using the bibtex program from suitable
% BiBTeX files (here: strings, refs, manuals). The IEEEbib.bst bibliography
% style file from IEEE produces unsorted bibliography list.

% -------------------------------------------------------------------------
\bibliographystyle{IEEEbib}
\bibliography{paper}
